Diversification of Australian blind snakes

### 2022-11-30 manuscript.Rmd
added some limitation statements and listed questions

In this study, we use the Australian radiation of blindsnakes as a system to investigate the diversification of morphologically conserved squamates by testing hypotheses about the processes that promote speciation.

@@

We 
- Estimate the ancestral biome of Australian blindsnakes based on extant taxa.
- Compare fit_env models between localised versus globalised temperature on the diversification of Australian blindsnake radiation, 
- Assess model fit between paleo-aridity and paleo-climate on diversification rates, and 
- Are there differences in the rates of speciation between arid and non-arid zones?

@@

## Biogeographic-dependent speciation

@@

## Reconstruction of Australian paleo climate
<!-- Consult Alex Skeels -->
To estimate paleoclimate, we used three paleographic digital elevation models (DEMs) from @scoteseAtlas2021, @straumeGlobal2020, and [VALDES citation]. Each DEM provide paleotopography at resolution of 1&deg; x 1&deg; and 5 myr resolution. We projected the paleoenvironmental reconstructions into an equal-area Behrmann projection of 110km x 110km following methods described in @hagenGen3sis2021. From the projected data we extracted temperature and aridity values across all grid cells that fall within Sahul by manually drawing a polygon that encompassed the continent from 40 mya to the present-day. Because reptiles are sensitive to extremely low and high temperature, we decided to to summarise the reconstructed paleo temperature data into three different data sets: minimum temperature, maximum temperature, and mean temperature.

@@

## Time ane environment-dependent diversification analyses

@@

## Ancestral biome reconstruction and biogeographic history

@@

Based on 50 BioGeoBEARS biogeographic stochastic mapping of our best model, while we inferred the tropical grassland biome to be the ancestral biome of Australian _Anilios_ (\@ref(fig:fig-biogeobears)), the ancestor leading to crown *Anilios* is ambiguous. We estimated an average of 71.2 ± 1.3 (mean ± SD) total anagenetic and cladogenetic events occured throughout the phylogeny, out of which the majority were a combination of anagenetic dispersal (31.2 ± 1.3 events) and founder events (6.1 ± 1.5 events) (see supp-summary_counts_bsm_table). 
We found that the largest source of outward dispersal is the tropical grassland biome (16.2 ± 4.9 events; 43.6%) making up almost double the number of outward dispersals from the arid biome (9.7 ± 3.2 events; 26.0%) (\@ref(fig:fig-biogeobears))c-d. Together the tropical grassland biome and the arid biome make up the largest source of diversificttion accounting for 69.6% of outward transitions.

On the other hand, the largest sink for all inward dispersal are the temperate grassland biome (7.9 ± 4.2 events; 21.2%) and the mediterranean biome (7.0 ± 2.9 events; 18.8%). The largest directional transition is from the arid biome to the mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland into the temperate forest biome (5.1 ± 1.3 events) (\@ref(fig:fig-biogeobears)c-d.

@@

We applied GeoHiSSE models to evaluate whether speciation, extinction, and turnover rates differ between lineages in arid, non-arid, and widespread ranges. We found that the model where variation in diversification rates were not linked to geographic range fit the data better than models where diversification rates were depdendent on geographic range (see supp-table-geohisse). The best-fitting geographic range independent model included a hidden unobserved state, indicating diversification rates is better explained by a hidden unobserved state. Compared to the three other models ,the best-fitting model accounts for 0.95 of the AIC weight and had AIC value 7.02 units lower than that of the second best-fitting model (see supp-table-geohiss). The weighted average parameter estimates from all models indicate both diversification rates and turnover rates did not differ between arid and non-arid ranges (REF FIGURE). However, net diversification rates for the widespread lineages were slightly higher (see figure), but this may be due to low amount of lineages.

@@

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 MYA, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021]. Paleoclimate estimates for Australia differed between the three methods (\@ref(fig:fig-paleoclimate-compare))b. However, all three paleo temperature models indicates that Australia was hotter than the average global temperature. Paleoclimate in Australia were also relatively constant compared to global temperature that shows greater fluctuation then slight decrease towards the present.

@@

We found that the exponential environment-dependent speciation and constant extinction (BEnvDCST) with aridity index derived from @scoteseAtlas2021 provided the best fit to the phylogeny (AICc = 238.43; see Table-RPANDA). This model indicates that increase aridity is exponentially associate with decrease speciation rate (lambda = 1.32, alpha = -0.14) and with a constant background extinction (mu = 0.23). The second and third most competitive models (delta AICc < 2) also found increase aridity is exponentially associated with decrease speciation rate but without estimating extinction (Table-RPANDA-TOP10). Among exponential environment-dependent models, we found Australia-specific paleoclimate data provides a better fit to our phylogeny than global paleoclimate (see supp-TABLE_FULL_RPANDA_model_fit).

## Rates of diversification

@@

- Limitations is we can only coarsely assign species to one biome one some are range-limited endemics on islands (e.g. _A. longissimus_)

@@

## Drivers of diversifcation 
- We expected the biogeographic models with jump dispersal to be the best fitting model because _Anilios_ are believed to have non-Australian origins [@marinTracing2013;@vidalBlindsnake2010;@oliverPhylogenetic2017].
- Blind snakes are thought to not move very much and so formation of geological barriers during aridification may accelerate speciation between 20 – 10 MYA. 
- The fact that DEC+J+x model was preferred over DEC+J may be due to poor dispersal ability as closely related species occupy the same or nearby areas. Events like floods are likely to have faciliated dispersals but in short distances. 
Discuss species that are far apart...
- Speciation could even be sympatric but no evidence to support this.

- High rates of diversification observed in concert with changing climate may be explained by allopatric speciation.

## Environment-dependent diversification
- Diversification rate decrease with aridification. 
- Our RPANDA model did not model environment-dependent extinction. 
- Extinction is a big cavaet especially because we suspect there are more species
-

@@

Account for heterogeneity of the arid zone [@pepperLife2021].

>>>>

@@

In this study, we use the Australian radiation of blindsnakes as a system to investigate the diversification of morphologically conserved squamates by testing hypotheses about the processes that promote speciation. Specifically, we sought to answer the following questions:
(1) What is the ancestral biome of Australian _Anilios_? 
() Which biomes are the largest source and sinks of biodiversity? 
() What is the relationship between diversification rate and changes in paleo environment? 
() Which paleo environment best explain current phylogeny?
() Are there differences in diversification rates between lineages in arid and non-arid biome?

Aims: 
- Infer the ancestral biome of Australian blindsnakes based on near-complete phylogeny of extant taxa.
- Investigate how paleo-environmental changes influence rates of diversification. 
- Assess model fit between paleo-environment.  
- Compare the rates of diversification between lineages in the arid versus non-arid biome.

- Compare localised versus globalised temperature on the diversification of Australian blindsnake radiation,

@@

## Biogeographic-dependent diversification analyses

@@

## Reconstruction of Australian paleo environment
<!-- Consult Alex Skeels -->
To estimate paleoclimate, we used three paleographic digital elevation models (DEMs) from @scoteseAtlas2021, @straumeGlobal2020, and [VALDES citation]. Each DEM provide paleotopography at resolution of 1&deg; x 1&deg; and 5 myr resolution. We projected the paleoenvironmental reconstructions into an equal-area Behrmann projection of 110km x 110km following methods described in @hagenGen3sis2021. From the projected data we extracted temperature and aridity values across all grid cells that fall within Sahul by manually drawing a polygon that encompassed the continent from 40 mya to the present-day. Because reptiles are sensitive to extremely low and high temperature, we decided to to summarise the reconstructed paleo temperature data into three different data sets: minimum temperature, maximum temperature, and mean temperature.

@@

## Environment-dependent diversification analyses

@@

## Ancestral biome reconstruction and transitions

@@

Based on 50 BioGeoBEARS biogeographic stochastic mapping of our best model, while we inferred the tropical grassland biome to be the ancestral biome of Australian _Anilios_ (\@ref(fig:fig-biogeobears)), the ancestor leading to crown *Anilios* is ambiguous. We estimated an average of 71.2 ± 1.3 (mean ± SD) total anagenetic and cladogenetic events occured throughout the phylogeny, out of which the majority were a combination of anagenetic dispersal (31.2 ± 1.3 events) and founder events (6.1 ± 1.5 events) (see supp-summary_counts_bsm_table).

We found that the largest source of outward dispersal is the tropical grassland biome (16.2 ± 4.9 events; 43.6%) making up almost double the number of outward dispersals from the arid biome (9.7 ± 3.2 events; 26.0%) (\@ref(fig:fig-biogeobears))c-d. Together the tropical grassland biome and the arid biome make up the largest source of diversificttion accounting for 69.6% of outward transitions. On the other hand, the largest sink for all inward dispersal are the temperate grassland biome (7.9 ± 4.2 events; 21.2%) and the mediterranean biome (7.0 ± 2.9 events; 18.8%). The greatest directional transition is from the arid biome to the mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland into the temperate forest biome (5.1 ± 1.3 events) (\@ref(fig:fig-biogeobears)c-d.

@@

We applied GeoHiSSE models to evaluate whether speciation, extinction, and turnover rates differ between lineages in arid, non-arid, and widespread ranges. We found that the model where variation in diversification rates were not linked to geographic range fit the data better than models where diversification rates were depdendent on geographic range (see supp-table-geohisse). The best-fitting geographic range independent model included a hidden unobserved state, indicating diversification rates is better explained by a hidden unobserved state. Compared to the three other models, the best-fitting model accounts for 0.95 of the AIC weight and had AIC value 7.02 units lower than that of the second best-fitting model (see supp-table-geohiss). The weighted average parameter estimates from all models indicate both diversification rates and turnover rates did not differ between arid and non-arid ranges (REF FIGURE). However, net diversification rates for the widespread lineages were slightly higher (see figure), but this may be due to low amount of lineages.

@@

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 MYA, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021]. Paleoclimate estimates for Australia differed between the three methods (\@ref(fig:fig-paleoclimate-compare))b. However, all three paleo temperature models indicate Australia was hotter than the average global temperature. Paleoclimate in Australia were also relatively constant (i.e. less fluctuation) compared to global temperature.

@@

We found that the exponential environment-dependent speciation and constant extinction (BEnvDCST) with aridity index derived from @scoteseAtlas2021 provided the best fit to the phylogeny (AICc = 238.43; see Table-RPANDA). This best-fitting model indicates that increase aridity is exponentially associated with decrease speciation rate (lambda = 1.32, alpha = -0.14) and a constant background extinction (mu = 0.23). The second and third most competitive models (delta AICc < 2) also found increase aridity is exponentially associated with decrease speciation rate but without estimating extinction (Table-RPANDA-TOP10). Among exponential environment-dependent models, we found Australia-specific paleoclimate data provides a better fit to our phylogeny than global paleoclimate (see supp-TABLE_FULL_RPANDA_model_fit).

@@

@@

## Drivers of diversifcation 
- We expected the biogeographic models with jump dispersal to be the best fitting model because _Anilios_ are believed to have non-Australian origins [@marinTracing2013;@vidalBlindsnake2010;@oliverPhylogenetic2017].
- Blind snakes are thought to not move very much and so formation of geological barriers during aridification may accelerate speciation between 20 – 10 MYA. 
- The fact that DEC+J+x model was preferred over DEC+J may be due to poor dispersal ability as closely related species occupy the same or nearby areas. Events like floods are likely to have faciliated dispersals but in short distances. 
Discuss species that are far apart...
- High rates of diversification observed in concert with changing climate may be explained by allopatric speciation. 
- Speciation could even be sympatric but no evidence to support this.

## Environment-dependent diversification
- Diversification rate decrease with aridification. But this is not explained by lineages transitioning into the arid biome as diversification rates did not differ between lineages in arid and non-arid biomes. 
- Our RPANDA model did not model environment-dependent extinction. 
- Extinction is a big cavaet especially because we suspect there are more species

## Limitations
There are several limitations in our models. 
- While taxonomic sampling the most representative of the genus, it is not comprehensive. We suspect there remains several undescribed species of blind snakes in Australia and New Guinea that may belong in the genus _Anilios_. Getting genetic data from _A. erycinus_ will allow us to place the species among other _Anilios_ and provide a more comprehensive picture of the evolution of this genus. 
- Limitations is we can only coarsely assign species to one biome one some are range-limited endemics on islands. For example *A. longissimus* has only been collected from 30m underground on Barrow Island located XX km the northwest coast of Western Australia. Placing that species with arid biome is not realistic but that area is considered the arid zone by our criteria.

- Need to account for heterogeneity of the arid zone [@pepperLife2021].

@@
